---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

**Overview**

The Behavioral Risk Factor Surveillance System (BRFSS) is an ongoing system designed to measure behavioral risk factors for the adult population (> 18 years old) in the US. It collects state data about U.S. residents regarding their health-related risk behaviors, chronic health conditions, and use of preventive services through telephone surveys. Initiated in 1984, BRFSS collected surveillance data on risk behaviors through monthly telephone interviews with 15 states. BRFSS now collects data in all 50 states as well as the District of Columbia and three U.S. territories. BRFSS completes more than 400,000 adult interviews each year.

**Survey Method**

The data is collected through telephone surveys. The surveys are conducted using Random Digit Dialing (RDD) techniques on both landlines and cell phones. RDD is a method for selecting people for involvement in telephone statistical surveys by generating telephone numbers at random. 

**Generalizability**

Regarding the survey method, data is collected through random sampling and hence can be generalized to the entire population. However, certain issues suggested potential bias of the results.

Assuming the US has a high telephone-ownership rate, when phone numbers are generated at random, each case in the population has an equal chance of being included and there is no implied connection between the cases in the sample. Hence, the data collected is regarded as random sample and it is believed to be representative of the population. Observational results from the sample can be generalized to the population.

Nevertheless, under a few circumstance, it is unclear whether the results are representative of the entire population. First, there would be bias introduced by non-responders. Non-response bias can be a problem if responders differ from non-responders for the measured variables. For example, non-responders may not have been contacted because they work multiple minimum-wage jobs. There would also be bias of voluntary responses, as people more interested in the subject tend to pay more attention and answer the questions. Further, it may occur that participants do not honestly answer some of the most sensitive questions about their health status, which carries the risk of a response bias.

**Causality**

Inference of a causal connection can only be done through experiment with random assignment. There is no evidence that random assignment is used with reference to the survey method. Hence, the data is not applicable to causal inference.

* * *

## Part 2: Research questions

**Research quesion 1:** 

Surveys found that women are more likely to have mental health problems than men, with young women at particularly high risk. In the society, there are often strong bias towards certain gender and age groups on mental issues. Are gender and age associated with mental health?

**Research quesion 2:**

Taking up exercise seems to reduce the risk of developing mental illness. It also seems to help in treating some mental health conditions. Does physical activities improve a person's mental health? How does it, in particular, associate with different psychological feelings like nervousness, hopelessness, restlessness, depression, hardship, and worthlessness?

**Research quesion 3:**

Relationship is considered to be correlated with health. Great marriage can bring better physical and psychological health. An unsatisfying marriage may raise stress and worsen our well-being. How relationship status is associated with both physical and mental health? Are there any difference of the association among different gender and age groups?


* * *

## Part 3: Exploratory data analysis

_Perform exploratory data analysis (EDA) that addresses each of the three research questions you outlined above. Your EDA should contain numerical summaries and visualizations. Each R output and plot should be accompanied by a brief interpretation._

**Research quesion 1:**


**Research quesion 2:**


**Research quesion 3:**

```{r}

```

