---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

**Describe how the observations in the sample are collected, and the implications of this data collection method on the scope of inference (generalizability / causality).**

### Overview

The Behavioral Risk Factor Surveillance System (BRFSS) is an ongoing system designed to measure behavioral risk factors for the adult population (> 18 years old) in the US. It collects state data about U.S. residents regarding their health-related risk behaviors, chronic health conditions, and use of preventive services through telephone surveys. Initiated in 1984, BRFSS collected surveillance data on risk behaviors through monthly telephone interviews with 15 states. BRFSS now collects data in all 50 states as well as the District of Columbia and three U.S. territories. BRFSS completes more than 400,000 adult interviews each year.

### Survey Method

The data is collected through telephone surveys. The surveys are conducted using Random Digit Dialing (RDD) techniques on both landlines and cell phones. RDD is a method for selecting people for involvement in telephone statistical surveys by generating telephone numbers at random. 

### Generalizability

Regarding the survey method, data is collected through random sampling and hence can be generalized to the entire population. However, if non-response rate is high, the results would be bias.

Assuming the US has a high telephone-ownership rate, when phone numbers are generated at random, each case in the population has an equal chance of being included and there is no implied connection between the cases in the sample. Hence, the data collected is regarded as random sample and it is believed to be representative of the population. Observational results from the sample can be generalized to the population.

Nevertheless, an important consideration in random digit dialing surveys is bias introduced by non-responders. Non-response bias can be a problem if responders differ from non-responders for the measured variables. For example, non-responders may not have been contacted because they work multiple minimum-wage jobs. Under this circumstance, it is unclear whether the results are representative of the entire population.

### Causality

Inference of a causal connection can only be done through experiment with random assignment. There is no evidence that random assignment is used with reference to the survey method. Hence, the data is not applicable to causation studies.

* * *

## Part 2: Research questions

**Come up with at least three research questions that you want to answer using these data. You should phrase your research questions in a way that matches up with the scope of inference your dataset allows for. Make sure that at least two of these questions involve at least three variables. You are welcomed to create new variables based on existing ones. With each question include a brief discussion (1-2 sentences) as to why this question is of interest to you and/or your audience.**

**Research quesion 1:**

**Research quesion 2:**

**Research quesion 3:**


* * *

## Part 3: Exploratory data analysis

**Perform exploratory data analysis (EDA) that addresses each of the three research questions you outlined above. Your EDA should contain numerical summaries and visualizations. Each R output and plot should be accompanied by a brief interpretation.**

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**

```{r}

```



**Research quesion 2:**

```{r}

```



**Research quesion 3:**

```{r}

```

