---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

*Describe how the observations in the sample are collected, and the implications of this data collection method on the scope of inference (generalizability / causality). Note that you might will need to look into documentation on the GSS to answer this question. See http://gss.norc.org/ as well as the "More information on the data" section below.*

* * *

## Part 2: Research question

*Come up with a research question that you want to answer using these data. You should phrase your research question in a way that matches up with the scope of inference your dataset allows for. You are welcomed to create new variables based on existing ones. Along with your research question include a brief discussion (1-2 sentences) as to why this question is of interest to you and/or your audience.*

* * *

## Part 3: Exploratory data analysis

*Perform exploratory data analysis (EDA) that addresses the research question you outlined above. Your EDA should contain numerical summaries and visualizations. Each R output and plot should be accompanied by a brief interpretation.*

* * *

## Part 4: Inference

*Perform inference that addresses the research question you outlined above. Each R output and plot should be accompanied by a brief interpretation.*
