---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

*Describe how the observations in the sample are collected, and the implications of this data collection method on the scope of inference (generalizability / causality). Note that you might will need to look into documentation on the GSS to answer this question. See http://gss.norc.org/ as well as the "More information on the data" section below.*

**Overview**

Since 1972, the General Social Survey (GSS) has been monitoring societal change and studying the growing complexity of American society. The GSS aims to gather data on contemporary American society in order to monitor and explain trends and constants in attitudes, behaviors, and attributes; to examine the structure and functioning of society in general as well as the role played by relevant subgroups; to compare the United States to other societies in order to place American society in comparative perspective and develop cross-national models of human society; and to make high-quality data easily accessible to scholars, students, policy makers, and others, with minimal cost and waiting.

GSS questions cover a diverse range of issues including national spending priorities, marijuana use, crime and punishment, race relations, quality of life, confidence in institutions, and sexual behavior.

**Study Method**

http://gss.norc.org/documents/codebook/GSS_Codebook_intro.pdf

The data come from the General Social Surveys, interviews administered to NORC national samples using a standard questionnaire. 

Each survey from 1972 to 2004 was an independently drawn
sample of English-speaking persons 18 years of age or over, living in non-institutional arrangements within the United States. Starting in
2006 Spanish-speakers were added to the target population. Block quota sampling was used in 1972, 1973, and 1974 surveys and for half of
the 1975 and 1976 surveys. Full probability sampling was employed in half of the 1975 and 1976 surveys and the 1977, 1978, 1980,
1982-1991, 1993-1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016, and 2018 surveys. Also, the 2004, 2006, 2008, 2010, 2012,
2014, 2016, and 2018 surveys had sub-sampled non-respondents.

**Generalizability**

**Causality**

* * *

## Part 2: Research question

*Come up with a research question that you want to answer using these data. You should phrase your research question in a way that matches up with the scope of inference your dataset allows for. You are welcomed to create new variables based on existing ones. Along with your research question include a brief discussion (1-2 sentences) as to why this question is of interest to you and/or your audience.*

* * *

## Part 3: Exploratory data analysis

*Perform exploratory data analysis (EDA) that addresses the research question you outlined above. Your EDA should contain numerical summaries and visualizations. Each R output and plot should be accompanied by a brief interpretation.*

* * *

## Part 4: Inference

*Perform inference that addresses the research question you outlined above. Each R output and plot should be accompanied by a brief interpretation.*
