---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

*Describe how the observations in the sample are collected, and the implications of this data collection method on the scope of inference (generalizability / causality). Note that you might will need to look into documentation on the GSS to answer this question. See http://gss.norc.org/ as well as the "More information on the data" section below.*

**Generalizability**

The survey data is collected through random sampling with different sampling methods like block quota sampling and full probability sampling. The sample size is less than 10% of the Unitied States population. Hence, the samples are considered independent of each other. Non-respondents are also sub-sampled since 2004, non-response bias is minimalized.

However, only English-speaking persons 18 years of age or over non-institutional arrangements within the United States are sampled from 1972 to 2004. Spanish-speakers were added to the target population since 2006. With non-English (and Spanish) speakers and institutional population excluded, inference within cannot be generalized to the entire United States population, but only the target population.

**Causality**

Inference of a causal connection can only be done through experiment with random assignment. There is no evidence that random assignment is used with reference to the survey method. Hence, the data is not applicable to causal inference.

* * *

## Part 2: Research question

*Come up with a research question that you want to answer using these data. You should phrase your research question in a way that matches up with the scope of inference your dataset allows for. You are welcomed to create new variables based on existing ones. Along with your research question include a brief discussion (1-2 sentences) as to why this question is of interest to you and/or your audience.*

* * *

## Part 3: Exploratory data analysis

*Perform exploratory data analysis (EDA) that addresses the research question you outlined above. Your EDA should contain numerical summaries and visualizations. Each R output and plot should be accompanied by a brief interpretation.*

* * *

## Part 4: Inference

*Perform inference that addresses the research question you outlined above. Each R output and plot should be accompanied by a brief interpretation.*
